---
title: "Election Data: <br> Demographic and Economic Variables <br> with Voting Outcomes"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'election.html'))})
author: "Jonni Johnson"
output: 
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

#Introduction


In this report, I include a summary report wherein data were documented at the county level of each state, and describe voter decisions broken down by demographic characteristics, economic trends, and past election patterns. Votes were cast in the 2016 Presidential Election. This report is broken down into separate sections:

+ *Data Prepping* - data merging/cleaning steps are described, research questions are identified, and new variables are constructed
+ *Data Analyses* - separated by research questions, R syntax is printed to outline analytical steps taken to arrive at each conclusion
+ *Data Summary with Visualization* - a general overview is provided as well as a short interpretation of each research question; results are presented in tables and as graphical depictions; additionally, corresponding code to create/manipulate data to accommodate ggplot2 specifications for graphing are included, where applicable

#Data Prepping

Loading data to prepare for merge
```{r}
cfacts = read.csv("county_facts (1).csv", sep=",", header = TRUE)
dim(cfacts)
head(cfacts)
subcfacts = cfacts[,names(cfacts) %in% c("fips", "area_name", "state_abbreviation", "SEX255214", "POP645213", "EDU635213", "EDU685213", "VET605213","INC110213", "PVY020213", "WTN220207", "RTN130207",  "POP060210")]

colnames(subcfacts) = c("fips", "county", "state", "femalep", "foreignpop", "highscho", "higherscho", "vet", "medhhinc", "povline", "wholesale", "retsale", "popdense")

subcfacts$state = droplevels(subcfacts$state)
subcfacts$county = droplevels(subcfacts$county)

eresults = read.csv("US_County_Level_Presidential_Results_12-16 (1).csv", 
                   sep = ",", header = TRUE)
eresults$state_abbr = droplevels(eresults$state_abbr)
eresults$county_name = droplevels(eresults$county_name)
dim(eresults)
names(eresults)

```

To merge the data, each data set must match on at least one column



We chose the column **fips** as both data sets had this same variable



##Cleaning/Merging
```{r}
colnames(eresults)[colnames(eresults) == "FIPS"] <- "fips"  #relabeled column name to merge
fullelect = merge(subcfacts, eresults, by = "fips")

str(fullelect)  #examine structure of dataframe 
```
Dropping some duplicate variables from our set (e.g., the "county_name" column so we don't use it by accident while doing analysis since it is from the eresults data set and is missing names for multiple counties).
```{r}
fullelect$county_name<-NULL
fullelect$X <-NULL 
fullelect$state <-NULL

str(fullelect)
```


###Research Questions

Several demographic and economic variables were considered: education level, presence of foreign born individuals, military status, gender majority by county, merchant and retail sales, percentage of indviduals below the poverty line, and median household income.


To answer whether the above variables were associated with voter outcome, multiple variables were constructed using Base R commands.


##Variable Creation {.tabset}

###Winner Variable
Creating a variable describing final voting outcome for each county:

Final Result (FR) of the election at each county:
```{r, echo=TRUE, warning=FALSE}
fullelect$FR<-""
fullelect$FR[fullelect$per_dem_2016 > fullelect$per_gop_2016] = "Dem"
fullelect$FR[fullelect$per_gop_2016 > fullelect$per_dem_2016] = "Gop"

FRtable <- aggregate(cbind(count = fullelect$FR) ~ fullelect$FR, data = fullelect, FUN = function(x){NROW(x)})

names(FRtable)[1] <- "County Outcome"

FRtable


```


*Coloring*

For this markdown, red palettes refer to Republican/GOP outcomes and blue palettes refer to Democratic:

````{r}


FRpalette <- c('navyblue', 'firebrick4') #creating a corresponding color palette to use later in graphing
FRpalletelite <- c("dodgerblue2", "firebrick1")

````



###Education Variable

Creating variables so that education levels can be divided into three categories
1. Percent without high school diploma
2. Percent with high school diploma but no higher level education
3. Percent with higher education (bachelor's or higher)

```{r}
#1
fullelect$without_high_school_diploma<-100-fullelect$highscho

#2
fullelect$high_school_education<-fullelect$highscho-fullelect$higherscho

#3
fullelect$higher_education<-fullelect$higherscho

colnames(fullelect)

summary(fullelect[,29:31])

#Assess that each adds to 100
edu.check<-fullelect$high_school_education+fullelect$higher_education+fullelect$without_high_school_diploma

summary(edu.check)

edu.check<-grepl(100,edu.check)
all(edu.check)


pct <- c(15.45, 64.8, 19.75)  #Graphing mean percentages of total voters
shades <- c("white", "lightgrey", "black")
categories <- c("No High School Diploma", "High School Diploma", "Higher Education Degree")



pie(pct, labels = categories, col = shades)



```

###Gender Majority Variable

Creating a variable to indicate the gender majority of each county
```{r}

fullelect <- cbind(fullelect, gendom = "female")

fullelect$gendom <- factor(fullelect$gendom, levels = c("female", "male"))

fullelect$gendom[fullelect$femalep < 50] <- "male" 

table(fullelect$gendom)
```
###Population Density in levels Variable

Re-classifying the population density variable to be in levels based on the national average, and above/below the 1st and 3rd quartiles.

1. First a summary statistics was called to identify level breaks

```{r}
summary(fullelect$popdense)
```
2. A new factor variable was constructed describing the county's density with regards to the national median (i.e., 45.2 people per sq mile)

```{r}
fullelect <- cbind(fullelect, dense = "")

fullelect$dense <- factor(fullelect$dense, levels = c("strongly less dense", "mildly less dense", "mildly more dense", "strongly more dense"))

fullelect$dense[fullelect$popdense < 16.9 ] <- "strongly less dense"

fullelect$dense[fullelect$popdense > 16.9 ] <- "mildly less dense"

fullelect$dense[fullelect$popdense > 45.2 ] <- "mildly more dense"

fullelect$dense[fullelect$popdense > 113.6 ] <- "strongly more dense"


table(fullelect$dense)
```



###Shifting Blue to Red Variable

Creating variables that consider how the county voted in 2012 compared to 2016, differences in percentage of vote for democratic party, at 5 levels of Democratic Vote changes from 2012-2016: Mild Blue Shift, Mild Red Shift, Moderate Red Shift, Strong Red Shift, Severe Red Shift

</b>

1. First a numeric difference variable was constructed (e.g., negative scores indicated greater Democratic percentage of the vote in 2016 compared to 2012)

2. Summary statistics defined the bounderies of each of the levels, mainly capturing the data around the mean and between the 1st and 3rd quartile of the data
```{r}
fullelect$demshift <- fullelect$per_dem_2012 - fullelect$per_dem_2016  # numeric shift variable

summary(fullelect$demshift)  #29 NA's due to missing 2012 data from ALASKA counties
```
  3. Create a categorical variable using the summary statistic boundaries
```{r}

fullelect <- cbind(fullelect, demmoreless = "")

fullelect$demmoreless <- factor(fullelect$demmoreless, levels = c("Mild Blue Shift", "Mild Red Shift", "Moderate Red Shift", "Strong Red Shift", "Severe Red Shift"))

fullelect$demmoreless[fullelect$demshift < 0 ] <- "Mild Blue Shift"

fullelect$demmoreless[fullelect$demshift > 0 ] <- "Mild Red Shift" #includes just beyond 1st quartile limit to include all data

fullelect$demmoreless[fullelect$demshift > 0.03353] <- "Moderate Red Shift"

fullelect$demmoreless[fullelect$demshift > 0.06788] <- "Strong Red Shift"

fullelect$demmoreless[fullelect$demshift > 0.10210 ] <- "Severe Red Shift"


str(fullelect$demmoreless)

table(fullelect$demmoreless)
```


# Data Analyses

##Demographic Questions

###Question 1
*Is there a relation between education level and election outcome?*

```{r}
edu<-aggregate(cbind(fullelect$higher_education,fullelect$high_school_education,fullelect$without_high_school_diploma)~fullelect$FR,FUN = "mean")
names(edu)[1]="Party"
names(edu)[2]="College Degree or Higher"
names(edu)[3]="High School Diploma Only"
names(edu)[4]="No High School Diploma"

edu
```
```{r}
library(ggplot2)
library(gridExtra) 
library(grid)
library(lattice)

ghighered <- ggplot(fullelect, aes(x = FR, y = higher_education)) +
           geom_boxplot(color = FRpalette, fill = FRpalletelite, width = .25) +
           scale_x_discrete(name = "College Degree") +
           scale_y_continuous(breaks=c(0, 20, 40, 60, 80), limits = c(0,90), name = NULL) +
           theme_classic()
           
ghighed <- ggplot(fullelect, aes(x = FR, y = high_school_education)) +
           geom_boxplot(color = FRpalette, fill = FRpalletelite, width = .25) +
           scale_x_discrete(name = "High School Diploma") +
           scale_y_continuous(breaks=c(0, 20, 40, 60, 80), limits = c(0,90), name = NULL) +
           theme_classic()

gnohighed <- ggplot(fullelect, aes(x = FR, y = without_high_school_diploma)) +
           geom_boxplot(color = FRpalette, fill = FRpalletelite, width = .25) +
           scale_x_discrete(name = "No High School Diploma") +
           scale_y_continuous(breaks=c(0, 20, 40, 60, 80), limits = c(0,90), name = NULL) +
            theme_classic()

title1=textGrob("Percentage of Voters by Educational Degree", gp=gpar(fontface="bold", fontsize = 18))

 grid.arrange(ghighered, ghighed, gnohighed, ncol = 3, nrow = 1, top =title1) 
```


| Dominant Party | College Degree or Higher   |  High School Diploma Only |  No High School Diploma|
|----------------|----------------------------|---------------------------|---------------------------|
| Democratic     | 27.95369                   | 56.36619                  | 15.68012                  |
| Republican     | 18.24542                   | 66.34926                  | 15.40531                  |

+ Greater percentage of people with higher education degrees appears to be associated with voting Democratic. Whereas the bulk of Republican swaying counties have greater percentages of High School Diploma Only or less.


###Question 2

*Is there a relation between percent of foreign born persons in a county and the election outcome?*

```{r}

foreign<-aggregate(fullelect$foreignpop~fullelect$FR,FUN = "mean")
names(foreign)[1]="Party"
names(foreign)[2]="Percentage of Foreign Born County Residents"
foreign
```
```{r}

summary(fullelect$foreignpop)


  
  ggplot(fullelect, aes(x = FR, y = foreignpop, color = FR)) +
               geom_violin(trim = T, fill = 'lightgrey') + 
    stat_summary(fun.data="mean_sdl", geom = "pointrange") +
    scale_color_manual(values = FRpalette) +
                 scale_y_continuous(name="Percentage of foreign born voters") +
               scale_x_discrete(name=NULL) +
               ggtitle("Distribution of Percentage Foreign Born Voters by Final Result") +
    theme_classic() +
    theme(legend.title =element_blank(), plot.title = element_text(hjust = .5)) +
    theme(legend.position = c(.1, .8))
    
       

```


| Dominant Party | Average % Foreign Born Personsin a county |  
|----------------|-------------------------------------------|
| Democratic     | 9.41168                                   |
| Republican     | 3.62009                                   |

+ Counties where the democratics won on average have higher percentages of foreign born persons. While both are skewed in distribution, many of the counties voting republican have less than 5% representation of foreign born citizens

```{r}
table(fullelect$FR, fullelect$foreignpop <= 5)

```

###Question 3 

*Is there a relation between number of veterans in a county and the election outcome?*

```{r, message=FALSE, warning=FALSE}
vetsum <- tapply(fullelect$vet, fullelect$FR, sum) 
vetsum  

library(scales)

vetsum3 <- tapply(fullelect$vet, list(fullelect$FR, fullelect$demmoreless), sum)
vetsum3 

fullvet <- ggplot(fullelect) +
           stat_summary(aes(x = FR, y = vet), fun.y=sum, geom = "bar", fill = c("royalblue2", "firebrick2"))
fullvet2 <- fullvet + 
            labs(x = NULL, y = NULL) +
            ggtitle("Total Number of Veterans by Each Party") +
            theme(plot.title =element_text(size = 16, face = "bold", hjust = 1),
             axis.text.x = element_text(size = 5),
             axis.text.y = element_text(size = 5),
             axis.title.x = element_text(size = 12),
             axis.title.y = element_text(size = 12)) +
             theme_classic() +
  theme(plot.title = element_text(hjust = .5)) + 
  scale_y_continuous(label = unit_format(unit = "m", scale = 1e-6))

fullvet2
```


| Dominant Party | Number of Veterans across all counties|  
|----------------|---------------------------------------|
| Democratic     | 9,700,890                             |
| Republican     | 11,562,245                            |


We might conclude that in general, veterans voted republican

But when we examine this while considering changing political opinions:


```{r, message=FALSE, warning=FALSE}
vet1 <- ggplot(na.omit(fullelect), aes(x = demmoreless, y = vet, fill= FR))

vet2 <- vet1 + geom_boxplot() + 
        scale_fill_manual(values = FRpalette) +
        scale_x_discrete(name = "") +
        scale_y_continuous(name = "Number of Veterans", limits = c(0, 30000), 
                          breaks = c(0,5000, 10000, 20000, 30000)) +
        labs(title = "Veteran Distribution by Shifting Opinion")
        

 vet2 + theme_classic() +
         theme(legend.title = element_blank()) +
   theme(plot.title = element_text(hjust = .5))


```

```{r}
ggplot(na.omit(fullelect), aes(x = demshift, y = vet/1000, color = FR)) + geom_point(alpha = .4) +
 facet_wrap(~dense, scales = "free_y") +
            scale_color_manual(values = c("royalblue2", "firebrick2")) +
            labs(x = "Conservative Shift", y = "Number of Veterans in Thousands") +
            ggtitle("Veterans by Conservative Shifts") +
            theme_classic() +
            theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) +
            theme(legend.title = element_blank())
             
```

| Dominant Party | Mild Blue Shift| Mild Red Shift| Moderate Red Shift| Strong Red Shift| Severe Red Shift|  
|----------------|----------------|---------------|-------------------|-----------------|-----------------|
| Democratic     |  4,598,673     |  3,095,515    |  1,210,601        |  625,142        |  170,959        |
| Republican     |  1,403,229     |  2,798,007    |  3,319,783        |  2,183,694      |  1,786,528      |


+ It appears that, when counties were leaning more democratic or experienced the weakest shift from blue to red, more veterans voted democratic; however, when counties were leaning more republican, greater number of veterans were observed among counties who went Republican. This seemed to be related as well to the population density of the area. Wherein, areas with mild blue or mild red shifts in opinion had greater veterans voting democrat than republican, particularly in densely populated areas


##Economic Questions

Questions 4 and 5 considered Merchant and Retail sales, respectively. Results are presented together for bigger picture take-away

###Question 4

*Is there a relation between value of merchant wholesaler county sales and the election outcome?*
Table Syntax
```{r}
A=aggregate(fullelect$wholesale~fullelect$FR,FUN = "mean")
names(A)[1]="Party"
names(A)[2]="Mean Wholesale Value (1000$)"
A


A1 <- tapply(fullelect$wholesale, list(fullelect$FR, fullelect$dense), mean)

A1
```
Graphing Syntax
```{r}
Aplot <- ggplot(fullelect) +
           stat_summary(aes(x = FR, y = wholesale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) + 
            labs(x = "", y = "Amount of Merchant Wholesales in Millions") +
            ggtitle("Merchant Wholesales") +
            theme_classic() +
            theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) +
            scale_y_continuous(label = unit_format(unit = "m", scale = 1e-6)) 


             
```


+ Considering Density of Area

```{r}
tmpAleastdense <- fullelect[fullelect$dense =="strongly less dense",]
Aleast <- ggplot(na.omit(tmpAleastdense)) +
           stat_summary(aes(x = FR, y = wholesale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Strongly Less", y = "Amount of Merchant Wholesales in Millions") +
            theme_classic() + 
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) 
 

tmpAlessdense <- fullelect[fullelect$dense =="mildly less dense",]

Aless <- ggplot(na.omit(tmpAlessdense)) +
           stat_summary(aes(x = FR, y = wholesale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Mildly Less", y = NULL) +
              theme_classic() +
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) 
  

tmpAmoredense <- fullelect[fullelect$dense =="mildly more dense",]
Amore <- ggplot(na.omit(tmpAmoredense)) +
           stat_summary(aes(x = FR, y = wholesale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Mildly More", y = NULL) +
              theme_classic() +
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) 
  


tmpAmostdense <- fullelect[fullelect$dense =="strongly more dense",]

Amost <- ggplot(na.omit(tmpAmostdense)) +
           stat_summary(aes(x = FR, y = wholesale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Strongly More", y = NULL) +
              theme_classic() + 
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) 
  

```


###Question 5

*Is there a relation between value of retail county sales and the election outcome?*
```{r}
B=aggregate(fullelect$retsale~fullelect$FR,FUN = "mean")
names(B)[1]="Party"
names(B)[2]="Mean Retail Sales Value (1000$)"
B

B1 <- tapply(fullelect$retsale , list(fullelect$FR, fullelect$dense), mean)
B1
```

```{r}
Bplot <- ggplot(fullelect) +
           stat_summary(aes(x = FR, y = retsale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2"))
Bplot2 <- Bplot + 
            labs(x = "", y = "Amount of Retail Sales") +
            ggtitle("Merchant Retail Sales") +
            theme_classic() + 
            theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) +
            scale_y_continuous(label = unit_format(unit = "m", scale = 1e-6))
```

+ Considering areas' density here as well

```{r}

Bleast <- ggplot(na.omit(tmpAleastdense)) +
           stat_summary(aes(x = FR, y = retsale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Strongly Less", y = "Amount of Merchant Retail Sales") +
              theme_classic() +
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) 


Bless <- ggplot(na.omit(tmpAlessdense)) +
           stat_summary(aes(x = FR, y = retsale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Mildly Less", y = NULL) +
              theme_classic() +
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) 


Bmore <- ggplot(na.omit(tmpAmoredense)) +
           stat_summary(aes(x = FR, y = retsale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Mildly More", y = NULL) +
              theme_classic() +
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5))  


Bmost <- ggplot(na.omit(tmpAmostdense)) +
           stat_summary(aes(x = FR, y = retsale), fun.y=mean, geom = "bar", fill = c("royalblue2", "firebrick2")) +
             labs(x = "Strongly More", y = NULL) +
             theme_classic() +
             theme(plot.title =element_text(size = 16, face = "bold", hjust = .5))  
```
####*Is there a relation between a county's commerce and the election outcome?*

| Dominant Party | Whole Sale   | Retail Sale |  
|----------------|--------------|-------------|
| Democratic     | 5,662,055.90 | 4,554,350   |
| Republican     | 392,957.80   | 637,637     |

```{r}
title2=textGrob("Commerce and Voting Behavior", gp=gpar(fontface="bold", fontsize = 18))

 grid.arrange(Aplot, Bplot2, ncol = 2, nrow = 1, top =title2)
```

+ As it can be seen, the counties with higher level of whole and retail sale values tend to vote Democratic while those with less whole and retail sale portions tend to vote Republican.

+ Considering density, counties leaned Repulican when of lower density and doing better in Whole sale compared to Democratic leaning counties at the same density level. Yet as denisty increased, so did trade as did county tendency to vote Democratic.

Whole Sale

| Dominant Party | Strongly less dense| Mildly less dense| Mildly more dense| Strongly more dense|  
|----------------|--------------------|------------------|------------------|--------------------|
|Democratic      |  26,379.97         |  79,090.65       |  744,395.30      |  10,041,393        |
|Republican      |  55,112.80         |  94,683.56       |  146,693.90      |  1,611,604         |
```{r}
title3=textGrob("Merchant Wholesales, Density, and Voting Behavior", gp=gpar(fontface="bold", fontsize = 14))

 grid.arrange(Aleast, Aless, Amore, Amost, ncol = 4, nrow = 1, top =title3)
```


Retail Sale

| Dominant Party | Strongly less dense| Mildly less dense| Mildly more dense| Strongly more dense|  
|----------------|--------------------|------------------|------------------|--------------------|
|Democratic      |  145,828.68        |  373,378.60      |  1,359,407.80    |  7,750,199.00      |
|Republican      |  96,647.82         |  208,089.20      |  461,970.60      |  2,221,097.00      |


+ Similarly in retail sales, when considering the density of the county compared to the national density average, as density increases, the disparity in the amount of retail sales between Democratic and Republican leaning counties increases.
```{r}
title4=textGrob("Merchant Retail Sales, Density, and Voting Behavior", gp=gpar(fontface="bold", fontsize = 14))

 grid.arrange(Bleast, Bless, Bmore, Bmost, ncol = 4, nrow = 1, top =title4)
```

+ Overall, counties exhibiting greater economic hardship with lower levels of retail sale tended to vote Republican.


###Question 6 

*Is there a relation between percentage of people living below the poverty line, gender majority, and election outcome?*

```{r}
pov <- tapply(fullelect$povline, list(fullelect$FR, fullelect$gendom), mean)
pov
```

```{r}
povplot <- ggplot(fullelect, aes(x = FR, y = povline)) +
            geom_boxplot(fill = FRpalletelite, color = FRpalette) +
            labs(x = NULL, y = "Percentage Below Poverty Line") +
             ggtitle("Poverty Level and Voting Behavior") +
             theme_classic() +
            theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) +
              theme(legend.title = element_blank())
povplot 

```

+ Counties that went Democratic tended to have a higher percentage of people living below the poverty line


| Dominant Party | Female Majority| Male Majority|  
|----------------|----------------|--------------|
|Democratic      | 19.10701       | 19.82479     |
|Republican      | 16.56970       | 15.59543     |

```{r}
povplot2 <- ggplot(fullelect, aes(x = per_dem_2016, y = povline, color = gendom)) +
            geom_point(size = 3, alpha = 0.6) +
            scale_color_manual(values = c("deeppink2", "deepskyblue2"))

povplot3 <-  povplot2 +  labs(x = "Percentage of Dem Vote", y = "Percentage Below Poverty Line") +
             #scale_color_manual(values = c("darkorchid3", "chocolate3")) +
             ggtitle("Poverty, Gender Majority, and Democratic Voting") +
              theme_classic() +           
              theme(plot.title =element_text(size = 16, face = "bold", hjust = .5)) +
              theme(legend.title = element_blank())

povplot3

```

+ The gender majority did not seem to affect the influence poverty percentage had on voter outcome


###Question 7

*Is there a relation between median income amount and changing political opinions?*

```{r}
medinc <- tapply(fullelect$medhhinc, fullelect$demmoreless, mean)
medinc

#Considering Gender Majority 

medincgen <- tapply(fullelect$medhhinc, list(fullelect$demmoreless, fullelect$gendom), mean)
medincgen

```

| Political Opinion Change | Female Majority| Male Majority|
|--------------------------|----------------|--------------|
|Mild Blue Shift           |  62,765.81     |  57,130.93   |
|Mild Red Shift            |  47,889.97     |  49,476.61   |
|Moderate Red Shift        |  43,820.07     |  43,043.92   |
|Strong Red Shift          |  43,288.16     |  45,094.94   |
|Severe Red Shift          |  42,971.00     |  44,780.70   |

+ Counties that increased in percentage of Democratic votes from 2012-2016 exhibited higher median incomes, particularly in female majority counties.

+ The median income is $44,160.00; unsupringly, counties where the average median income is below the national average show an increase in shifting away from Democratic votes and toward voting Republican

+ It is possible that the shift observed in percentage of Democratic votes is from frustration at stale economic growth; whereas counties that are well above the average show a preference to permit Democratics to stay in control.


#Data Summary


Demographic characteristics, economic trends, and prior voting behavior were considered in the analyses reported. Our analyses have come to four broad conclusions regarding who voted Republican versus Democratic:

1. Arguably, less exposure to diversity (i.e., by way of foreign born neighbor or higher education) was associated with greater tendency to vote Republican

2. Veterans heavily contributed to the county vote going republican, when the county was moderately shifting (or greater) in favor of Republican. This however was not the case in densely populated areas; here, the trend was reversed with more veterans populating democratic leaning counties

3. Counties, particularly as population density increased, in which economic conditions of the community have been better in terms of commerce tend to vote Democratic.

4. The socio-economic analyses together seem to indicate that counties, particularly female majority counties, that were better off (i.e., higher median income) or had greater percentage of individuals below the poverty line tended to vote democratic. It appears as those the individual in the middle, neither experience too much hardship or any economic growth, exhibit a shift in voting preference, prefering the Republican party, especially in lesser, densly populated areas.

